
Programming Assignment #1
=========================
Jared Kramer


```{r setup, echo = TRUE}
knitr::opts_chunk$set(echo = TRUE)
```


### 1. Loading and Processing the Data

First, unzip the activity data if necessary, and read in the csv file into a data frame. 
To allow for splitting by date for later analysis, we convert the date field into factors. 

```{r getdata}
  if(!file.exists("activity.csv")){
    unzip("activity.zip")
  }

  activityDF <- read.csv("activity.csv")
  activityDF$date <- as.Date(activityDF$date)
  
```


### 2. What is the mean total of steps taken per day?

In this section, we split the "steps" column of the data frame, according to the date of each observation.  

```{r dailysteps}
  steps_by_day <- split(activityDF$steps, activityDF$date)
  dailysteps <- sapply(steps_by_day, sum)
  hist(dailysteps, breaks = 20, main = "Histogram of Steps per Day", xlab = "Daily number of Steps")
```

Ignoring NA's, we then compute the mean and the median, and display it inline below.

```{r meanmedian}
  mean1 <- mean(dailysteps, na.rm = TRUE)
  median1 <- median(dailysteps, na.rm = TRUE)
```

The mean is `r as.character(mean1)` and the median is `r as.character(median1)`.

### 3. What is the average daily activity pattern? 

In this section, we split the "steps" column of the data frame, according to the time interval of each observation. 

```{r intervalsteps}
  steps_by_interval <- split(activityDF$steps, activityDF$interval)
  intervalsteps <- sapply(steps_by_interval, mean, na.rm = TRUE)
  
  max_interval <- names(intervalsteps)[which.max(intervalsteps)]
  
  plot(x = names(intervalsteps), y = intervalsteps, type = "l", 
       xlab = "Average Steps in Time Interval", 
       ylab = "Time Interval")
```

The interval with the maximum average number of steps per day is `r max_interval`


